%\documentclass[typewriterstyle]{thesis}
\documentclass[defaultstyle,11pt]{thesis}
%\documentclass[modernstyle,11pt]{thesis}
%\documentclass[defaultstyle,11pt,twoside,openright]{thesis}  %%  double-sided 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%  see documentation for information about  %%%%%%%%%%
%%%%%%%%%%%  the options (11pt, defaultstyle, etc.)   %%%%%%%%%%
%%%%%%%  http://www.colorado.edu/its/docs/latex/thesis/  %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%		\documentclass[typewriterstyle]{thesis}
% 		\documentclass[modernstyle]{thesis}
% 		\documentclass[modernstyle,11pt]{thesis}
%	 	\documentclass[modernstyle,12pt]{thesis}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%    load any packages which are needed    %%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{latexsym}		% to get LASY symbols
\usepackage{graphicx}		% to insert PostScript figures
\usepackage{rotating}		% for sideways tables/figures
\usepackage{natbib}
\usepackage{deluxetable}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{hyperref}  % not working yet

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%       all the preamble material:       %%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use ApJ style citations in text
\bibpunct{(}{)}{;}{a}{}{,}

%\renewcommand{\bibstyle@aas}{\bibpunct{(}{)}{;}{a}{}{,}}% 
%\@citestyle{aas}% 


\title{Convection and Dynamo Action in Rapidly Rotating Suns}

\author{B.~P.}{Brown}

\otherdegrees{B.S., Harvey Mudd College, 2003}

\degree{Doctor of Philosophy}		%  #1 {long descr.}
	{Ph.D., Astrophysical and Planetary Sciences}   %  #2 {short descr.}

\dept{Department of}			%  #1 {designation}
	{Astrophysical and Planetary Sciences}		%  #2 {name}

\advisor{Prof.}				%  #1 {title}
	{Juri Toomre}			%  #2 {name}

\reader{John Bally}
\readerThree{Allan Sacha Brun}
\readerFour{Keith Julien}         %  4rd person to sign thesis
\readerFive{Mark Miesch}		%  3rd person to sign thesis
\readerSix{Mark Rast}		%  2nd person to sign thesis


%\readerSix{Prof.~Michael Shull}
%\readerSeven{Prof.~Allen Sacha Brun}

\abstract{  %\OnePageChapter	% one page only ??

    When our Sun was young it rotated much more rapidly than 
it does now. Observations of young, rapidly rotating stars indicate that many
possess substantial magnetic activity and strong axisymmetric magnetic
fields.  There is furthermore an observed correlation between the
stellar rotation rate and surface magnetism.  Yet the origins of the
magnetic activity or the correlation with rotation remain unclear.
We conduct simulations of dynamo action in rapidly
rotating suns with the 3-D MHD anelastic spherical harmonic (ASH) code
to explore the complex coupling between rotation, convection and
magnetism.   Here we study global-scale flows of differential rotation
and meridional circulation as well as dynamo action realized in the
bulk of the convection zone for stars rotating from one to fifteen
times the current solar rate. 

We find that more rapidly rotating stars generally have stronger
flows of differential rotation but weaker meridional circulations that
break into multiple cells in both radius and latitude.  Surprising
localized states arise in the rapidly rotating simulations, with
convection modulated in longitude.  In the most rapid rotators
convection can be entirely confined to narrow active nests which
persist for thousands of days and propagate through the shearing flow
of differential rotation at their own distinct velocity.

We find that substantial organized global-scale magnetic fields are
achieved by dynamo action in these rapidly rotating suns. Striking wreaths of
magnetism are built in the midst of the convection zone, coexisting
with the turbulent convection. This is a great surprise, for
many solar dynamo theories have suggested that a tachocline of
penetration and shear at the base of the convection zone is a crucial
ingredient for organized dynamo action, whereas these simulations
generally do not include such tachoclines. Some dynamos achieved in these rapidly
rotating states build persistent global-scale fields which maintain
amplitude and polarity for thousands of days.  Other dynamos can
undergo cycles of activity, with fields varying in strength and even
changing in global-scale polarity.  As the
magnetic fields wax and wane in strength, the primary response in the
convective flows involves the axisymmetric differential rotation,
which begins to vary on similar time scales.  Bands of relatively fast
and slow fluid propagate toward the poles on time scales of roughly
500~days.  In the Sun, similar patterns are observed in the poleward
branch of the torsional oscillations, and these may represent a
response to poleward propagating magnetic field deep below the solar
surface.  

In one simulation, rotating at three times the solar rate,
we explore how the wreaths of magnetism are built and maintained by
the differential rotation and the turbulent correlations.  We further
explore whether a simple mean-field theory can reproduce our 3-D results
and find several discrepancies.  We generally find that
wreath-building dynamos are present in every region of parameter space
we have sampled, including simulations of the solar dynamo.  We find
that previous simulations had bottom boundary conditions which make
wreath formation difficult if not impossible, but that new simulations
of the solar dynamo can produce strong magnetic wreaths.
Lastly, we show that wreaths of magnetism survive in the presence of a
tachocline of penetration and shear at the base of the convection zone. 
These wreaths fill the convection zone and undergo quasi-regular
reversals of global-scale polarity.
 
}

\dedication[Dedication]{	% NEVER use \OnePageChapter here.
	To my mother and father who have encouraged every step along
        this path of discovery, and to Mindy who has made the journey one of joy.
	}

\acknowledgements{\vskip-0.5cm %\OnePageChapter	% *MUST* BE ONLY ONE PAGE!
This journey has only been possible with the help of many friends.
First and foremost, I thank my advisor Juri Toomre for his support,
friendship, generosity and wise advice.  Juri has provided
me with opportunities beyond compare. I cherish the many evenings I
have spent with Juri and his wife Linda, and look forward to the many
wonderful meals and celebrations to be shared together in the years to come.  

My collaborators and co-authors have become close friends over the
years.  Sacha Brun has provided constant and good advice, and
reflections on scientific careers and life, even as he has undertaken his
own journeys on these paths.  His European perspective is always
refreshing.  I have always looked forward with pleasure to my visits with
Mark Miesch at HAO.  His  positive outlook and deep insights have contributed greatly to my own
growth, and his excitement about the discoveries contained herein has
kept my own alive.  His thesis has been at my side for
many years, and it became particularly helpful in the past months; I
hope to one day write as well he does regularly.  And Mark, you're
right: the high mountains do soothe the soul.  My thanks also go to Matt
Browning, who walked these steps before me, and is now helping me to
find my own way.
Over the years Mark Rast has been a sage source of advice, helping me 
remember what really matters in the pursuit of scientific discovery,
and reminding me that the standard path to success is not the only one,
and that it is perhaps not always the the best.
Gwen Dickinson has been a friend and a constant eye of calm in the
whirling storm that can be our research.  She has remained
unflappably calm in the face of looming deadlines, and has saved us
all from disaster (myself in particular) many, many times.  

My office mates over the years, Jason Lisle, Geoff Vasil and Nick
Nelson, have each helped shape this continuing adventure.  Jason reminded me
early on of the importance of balancing life and science, and the
value of time.  Geoff has become a dear friend, and
many evenings have been spent together thinking on science and life,
in our tower office or over good food and good beer.  We both venture forth
now to postdocs in colder climes. Nick has helped me realize that I
really have finally learned a few things of use.  Good luck sir, on
your own path to discoveries!  Kyle Augustson and Nick Featherstone
are part of our cohort as well; I would not be here now with out their
aid.  I look forward to seeing where each of us ventures in the future.
Along my path, I've had the pleasure of meeting and working with John
Clyne and Alan Norton at NCAR, and Carter Emmart at the American
Museum of Natural History.  These three have helped me realize how
beautiful our science can be, and how enthusiastically it can be
received when shown in the right light.  With them, I look forward to
many fruitful years ahead.

My close friends and housemates, Brennan Gantner and Than Putzig, have
joined me in many wonderful adventures during our years in Boulder.
To them and Licia Ray, I owe my sanity.   Ben Mates and Dave
Schneider held the final watch at the Greenbriar estate, through the
craziness of moving, writing and planning a wedding; during these
times, they reminded me that this too would pass.  

These adventures in science began at a young age when my father told
me ``there are four fundamental forces of nature.  You should learn
about them, how they work, what they are.''  This fueled a
love for physics and then astronomy.  My mother and father have
supported me with love on every step of this adventure.
In recent days my path has been made clearer by the presence of my
beloved, Mindy Wilkinson, who has been my companion and guide.  
During dark days she reminds me that what we do is fundamentally
really, really neat.  That it matters.  That people care.  And to me, that
has value beyond words.  On the journeys ahead, we will tread our paths
together, which brings me incomparable joy.

I thank Juri Toomre, Mark Rast, Mark Miesch, Keith Julien, Sacha Brun,
and John Bally for serving on my thesis committee.  Over the years,
they each have acted to kindle my scientific enthusiasm, to provide
perspective on life, and to help me find my way.
It was an honor to have them present at the close of this chapter of life.

~\vspace{5cm}

This research has been supported by NASA through Heliophysics Theory
Program grants NNG05G124G and NNX08AI57G.  I have additionally been
supported through the NASA GSRP program by award number
NNG05GN08H.   The simulations in this thesis were carried out with NSF
PACI support of SDSC, PSC, TACC and NICS.
In particular, these simulations ran variously on
Blue Horizon, Datastar and Bluegene at SDSC, on Lemieux and Bigben at
PSC, on Ranger at TACC and on Kraken at NICS.  Volume renderings used in the analysis and the
field line tracings shown were produced using VAPOR
\citep{Clyne_et_al_2007}.  None of this analysis would have been
possible without the resources at Laboratory for Computational
Dynamics (LCD) here at the University of Colorado.


	}

\ToCisShort	% a 1-page Table of Contents ??

\LoFisShort	% a 1-page List of Figures ??
%	\emptyLoF	% no List of Figures at all ??

\LoTisShort	% a 1-page List of Tables ??
%	\emptyLoT	% no List of Tables at all ??


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%       BEGIN DOCUMENT...         %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%  footnote style; default=\arabic  (numbered 1,2,3...)
%%%%  others:  \roman, \Roman, \alph, \Alph, \fnsymbol
%	"\fnsymbol" uses asterisk, dagger, double-dagger, etc.
%	\renewcommand{\thefootnote}{\fnsymbol{footnote}}
%	\setcounter{footnote}{0}

\begin{document}

\input macros.tex       % file containing author's macro definitions
\input chapter_1.tex			% Introduction
\input chapter_2.tex			% Code description (MHD)
\input chapter_3.tex			% Convective patterns in hydro cases
\input chapter_4.tex			% Nests of convection, hydro
\input chapter_5.tex			% Dynamo case D3
\input chapter_6.tex			% Dynamo case D5
\input chapter_7.tex			% Production of field in dynamo case D3
\input chapter_8.tex			% Broader classes of dynamos
\input chapter_9.tex			% Directions for the future: Wreath building in the Sun

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%  Bibliography %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%\bibliographystyle{apj}
%\bibliographystyle{gvthesis}
%\bibliographystyle{bpbrown_thesis} % based on Geoff's file, but
 				   % designed to match ApJ format better
\bibliographystyle{bpbrown_thesis_apj} % based on apj.bst file
%\bibliographystyle{plain}	% or "siam", or "alpha", or "abbrv"
				% see other styles (.bst files) in
				% $TEXHOME/texmf/bibtex/bst

%\nocite{*}		% list all refs in database, cited or not.

\bibliography{bibliography}		% bib database file refs.bib

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%  Appendices %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\appendix	% don't forget this line if you have appendices!

%\input appendix_A.tex			% file with Appendix A contents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%   THE END   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

